The microbiota of the proximal and distal human colon

### 2017-09-05 submission/manuscript.Rmd
manuscript with tweaks for DKT review

#### Random Forest classification models identify important Operational Taxonomic Units (OTUs) on each side
To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models trained using OTU abundances. We used 10-fold cross validation to build the first model to classify the lumenal versus mucosal samples for the proximal and distal sides, independently (Figure 4A). The constructed model used `r length(2:ncol(aucrf_data_proximal_bs))` features for the proximal and `r length(2:ncol(aucrf_data_distal_bs))` for the distal. The models performed well when classifying these samples (`r signif(cv10f_roc_proximal_bs$auc, digits=3)` and `r signif(cv10f_roc_distal_bs$auc, digits=3)`, respectively). The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the proximal lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera (Figure 4C). These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.

Next, we built a model to differentiate the proximal and distal lumenal samples using 10-fold cross validation. The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, AUC = `r signif(cv10f_roc_muc$auc, digits=3)`) compared to the proximal versus distal lumen (AUC = `r signif(cv10f_roc_lum$auc, digits =3)`). The model that separated proximal and distal mucosal samples used `r length(2:ncol(aucrf_data_LRlumen))` features. OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B). The model used `r length(2:ncol(aucrf_data_LRbowel))` features to separate proximal and distal lumenal samples. Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C). This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also helped to differentiate between the two sides- such as _Anaerococcus_ and _Finegoldia_.

@@

To get around the noisiness from a diverse set of samples, we built a Random Forest model to identify microbes specific to each side. For each comparison we identified top 5 OTUs that were strongly predictive of one site or another. Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis @Donaldson2015. For instance, the proximal mucosa contains the highest oxygen concentratons of the colon and  harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and aerobic _Psuedomonas_. The distal mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_. The model was less effective at classifying the proximal and distal lumenal contents, probably because the samples are arguably composed of the same bacteria but differ in water content.

>>>>

#### Random Forest classification models identify important Operational Taxonomic Units (OTUs) on each side
To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models trained using OTU abundances. We used 10-fold cross validation to build the first model to classify the lumenal versus mucosal samples for the proximal and distal sides, independently (Figure S1A). The models were constructed to only use the five most predictive OTUs as input to reduce overfitting between samples. The models performed well when classifying these samples (`r signif(cv10f_roc_proximal_bs$auc, digits=3)` and `r signif(cv10f_roc_distal_bs$auc, digits=3)`, respectively). The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the proximal lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4A). The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera (Figure 4B). These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.

Next, we built a model to differentiate the proximal and distal lumenal samples using 10-fold cross validation. The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = `r signif(cv10f_roc_muc$auc, digits=3)`) compared to the proximal versus distal lumen (AUC = `r signif(cv10f_roc_lum$auc, digits =3)`). These models were also constructed to use only the five most predictive OTUs as input to reduce overfitting. OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5A). Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5B). This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also helped to differentiate between the two sides- such as _Anaerococcus_ and _Finegoldia_.

@@

To get around the noisiness from a diverse set of samples, we built a Random Forest model to identify microbes specific to each side. For each comparison we identified the top five OTUs that were strongly predictive of one site or another. Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis @Donaldson2015. For instance, the proximal mucosa contains the highest oxygen concentratons of the colon and  harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and aerobic _Psuedomonas_. The distal mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_. Thus the gut microenvironment of each location likely enriches for these specific microbes.

In addition to identifying features that are specific to each side of the gut, the ability of the Random Forest to classify samples can serve as a proxy for similarity. That is, a higher AUC value means the samples are more efficiently classfied (and thus more different) than a model with a lower AUC value. For instance, the model separating the proximal and distal mucosa has an AUC of `r signif(cv10f_roc_muc$auc, digits=3)` whereas the model for classifying the proximal and distal lumen has a much lower AUC of `r signif(cv10f_roc_lum$auc, digits =3)`. Further, prior to reducing the input to five OTUs, the latter model required 44 OTUs to best separate the samples. The much lower AUC and need for a high number of features compared to other models suggest these locations are the most similar of the comparisons tested. We speculate that the model was least effective at classifying the proximal and distal lumenal contents because the samples are arguably composed of the same bacteria but differ in water content.
